Learning and comparing functional connectomes across subjects

### 2013-01-01 paper.tex
iter

Predictive modeling is concerned with learning (or fitting) a model to
draw conclusions on new and unseen data \cite{pereira2009}. Typically on
brain imaging data, it opens the door to learning diagnosis or prognosis
markers \cite{stonnington2010}. However, beyond the practical use of
prediction, these models can also be useful to uncover mechanism of brain
function \cite{norman2006}. Both the fact that information \emph{can} be
drawn to generalize predictions to new data and \emph{how} or
\emph{where} informs us on brain function. Multivariate predictive models
are attractive in connectomics because they are sensitive to dependencies
between features and avoid the need to correct for multiple comparisons
since the significance of an entire pattern is evaluated using a single
statistical test. From the perspective of medical applications, using
resting-state connectivity as a biomarker is particularly appealing as it
can be applied to impaired patients \cite{greicius2008b}. Predictive
modeling techniques have been successfully applied to identify
connectome-based biomarkers of Alzheimer's disease
\cite{stonnington2010}, depression \cite{craddock2009}, schizophrenia
\cite{cecchi2009}, autism \cite{anderson2011}, ADHD \cite{zhu2008}, aging
\cite{dosenbach2010}, as well as to classify mental operations
\cite{richiardi2011,shirer2012}.

@@

Technically, predictive modeling is a supervised machine learning problem
where a target to be predicted --\emph{e.g.}\ age, disease state,
cognitive state-- is available for each observation of the data. In the
context of comparing connectomes, features used in the predictive model
usually correspond to bivariate measures of connectivity, or any other
summary metrics discussed earlier.
%
The prediction accuracy of a model for predicting the target variable
provides a natural means for evaluating the utility of the model as a
biomarker, but also its explanatory power for the mechanism under study.
%
This prediction accuracy is measured using one or more iterations of
cross validation, which involves training and testing phases
\cite{pereira2009}.  A cross validation cycle begins by separating the
available data into two independent subsets: training data and testing
data.  Training involves learning a predictive model from the training
data. During testing the model is applied to the testing data to predict
corresponding values of the target variable.  Prediction accuracy is then
assessed by comparing these predictions to the observed values of the
target variable.

% XXX: sentence below should probably be moved elsewhere
Support vector machines (classification and regression) are popular
algorithms for predictive modeling as they do not require fine parameter
setting.
% XXX: need better explanation, in particular of nested cross-val
However, learning algorithms require in general the specification of
parameters, the values of which will impact the quality of the results.
These parameters may be specified based on domain specific knowledge, or
requirements, determined using an analytical approach, or optimized using
cross validation. In latter case a nest cross-validation scheme is
required to avoid biasing resultant measures of prediction accuracy.

Determining the subset of features most relevant to the model (feature
selection) is essential for interpreting the models. In cases were model
weights can be directly mapped to their corresponding feature, there is
typically no statistical framework for thresholding these weights, and
meaningless features are not always set to zero (this is particularly the
case for $\ell_2$ regularized algorithms such as SVM and ridge
regression). One feature selection strategy is to filter out features
based on their statistical relationship with the variable of interest
using a univariate statistical criterion such as t-test, Pearson's
correlation, or mutual information.  A drawback of this approach is that
since the filters are typically insensitive to multivariate
information, they may exclude features that would be important to the
model.

% XXX: mention nested cross-val for feature selection above

@@

% XXX: distinguish linear from non-linear classifiers. Explain that a
% linear classifier is nothing more than a fancy linear regression

The interpretation of the predictive model can be complicated and
requires insight into the mechanisms
underlying the modeling approach. For linear classifiers, 
classifier weights are similar to the weights of a ordinary regression.
If features are appropriately
standardized prior to training, the magnitude of weights can be
interpreted to reflect the relative importance of the feature to the
model, but interpreting how the feature differs between classes or
relates to a continuous variable is more complicated, particularly given
the multivariate nature of their involvement.  For classification methods
such as Fisher's linear discriminant analysis (LDA), which maximally
separates class means given their standard deviation, feature weights are
representative of the differences between class means. For support vector
classification, classifier weights are derived from observations that are
the most similar between classes, and hence represents the border between
classes. To draw conclusions on the features used for prediction, it is
perhaps most reasonable to adapt a conservative interpretation in which
predictive models is used to identify the most relevant features that are
later tested in followon experiments better suited to elucidating their
relationship to the variable of interest.

>>>>

Predictive modeling is concerned with learning (or fitting) a model to
draw conclusions on new and unseen data \cite{pereira2009}. Typically on
brain imaging data, it opens the door to learning diagnosis or prognosis
markers \cite{stonnington2010}. However, beyond the practical use of
prediction, these models can also be useful to uncover mechanism of brain
function \cite{norman2006}. Both the fact that information \emph{can} be
drawn to generalize predictions to new data and \emph{how} or
\emph{where} it lies informs us on brain function. Multivariate
predictive models
are attractive in connectomics because they are sensitive to dependencies
between features and avoid the need to correct for multiple comparisons
since the significance of an entire pattern is evaluated using a single
statistical test. From the perspective of medical applications, using
resting-state connectivity as a biomarker is particularly appealing as it
can be applied to impaired patients \cite{greicius2008b}. Predictive
modeling techniques have been successfully applied to identify
connectome-based biomarkers of Alzheimer's disease
\cite{stonnington2010}, depression \cite{craddock2009}, schizophrenia
\cite{cecchi2009}, autism \cite{anderson2011}, ADHD \cite{zhu2008}, aging
\cite{dosenbach2010}, as well as to classify mental operations
\cite{richiardi2011,shirer2012}.

@@

Technically, predictive modeling is a supervised machine learning problem
where a target to be predicted --\emph{e.g.}\ age, disease state,
cognitive state-- is available for each observation of the data. In the
context of comparing connectomes, features used in the predictive model
usually correspond to bivariate measures of connectivity, or any other
summary metrics discussed earlier.
%
The prediction accuracy of a model for predicting the target variable
provides a natural means for evaluating the utility of the model as a
biomarker, but also its explanatory power for the mechanism under study.
%
This prediction accuracy is measured using one or more iterations of
cross validation, which involves training and testing phases
\cite{pereira2009}. Cross validation cycles separate the
available data into two independent subsets: training data and testing
data.  Training involves learning a predictive model from the training
data. During testing the model is applied to the testing data to predict
corresponding values of the target variable.  Prediction accuracy is then
assessed by comparing these predictions to the observed values of the
target variable.

Support vector machines\footnote{See \cite{kloppel2008}
for a gentle introduction to support vector machines.} (classification and regression) are popular
algorithms for predictive modeling as they work on a large class of problems 
and do not require fine parameter setting.
However, in general learning algorithms do require the specification of
parameters, the values of which will impact the quality of the results.
These parameters may be specified based on domain specific knowledge,
determined using an analytical approach, or set using
cross validation to optimize the prediction accuracy. In latter case 
to avoid biasing measures of prediction accuracy, a
nested cross-validation scheme \cite{pereira2009} is required, assessing 
prediction on left-out data not used to set the parameters.
%
Selecting features most relevant to the model can help prediction
accuracy and model interpretation \cite{pereira2009}. One feature selection strategy is 
to filter out features
based on their statistical relationship with the variable of interest
using a univariate statistical criterion such as t-test, Pearson's
correlation, or mutual information.  A drawback of this approach is that
since the filters are typically insensitive to multivariate
information, they may exclude features that would be important to the
model. Note that feature selection must be performed inside the
cross-validation loop, \emph{i.e.}\ on the train data, excluding the test
data.

@@

The interpretation of the predictive model can be complicated and
requires insight into the mechanisms
underlying the modeling approach. For linear classifiers, 
classifier weights are similar to the weights of a ordinary regression.
If features are appropriately
standardized prior to training, the magnitude of weights can be
interpreted to reflect the relative importance of the feature to the
model, but interpreting how the feature differs between classes or
relates to a continuous variable is more complicated, particularly given
the multivariate nature of their involvement\footnote{Note that there is
typically no statistical framework for thresholding these weights, and
meaningless features are not always set to zero (this is particularly the
case for $\ell_2$ regularized algorithms such as SVM and ridge
regression)}. For classification methods
such as Fisher's linear discriminant analysis (LDA), which maximally
separates class means given their standard deviation, feature weights are
representative of the differences between class means. For support vector
classification, classifier weights are derived from observations that are
the most similar between classes, and hence represents the border between
classes. To draw conclusions on the features used for prediction, it is
perhaps most reasonable to adapt a conservative interpretation in which
predictive models is used to identify the most relevant features that are
later tested in followon experiments better suited to elucidating their
relationship to the variable of interest.
